### 2018-04-19 1_intro.tex
Some editions. Rename output folder out -> build.

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{Project web page: \url{diy.inria.fr}} is a sofware suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy}, the litmus tests concrete executor \tool{litmus} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd} that implements reachability analysis for exploring states reachable under specified~WMM.

>>>>

Although weak memory studies is rather young research area, there exist frameworks and tools for exploring WMMs and examining simple programs with respect to the them. The state-of-the-art tool is \tool{diy} (for \textit{do it yourself}), developed by the researchers from INRIA institute, France and University of Cambridge, UK.
The \tool{diy}\footnote{Project web page: \url{diy.inria.fr}} is a sofware suite for designing and testing weak memory models. It is firstly released back in 2010, and since that time it remained to be the only tool for testing weak memory models. The \tool{diy} consists of several modules: the litmus tests generators \tool{diy}, \tool{diycross} and \tool{diyone}, the litmus tests concrete executor \tool{litmus} that runs tests on a physical machine while collecting its behaviours, and the weak memory models simulator \tool{herd} that implements reachability analysis for exploring states reachable under specified~WMM.

### 2018-04-19 2_wmm.tex
Some editions. Rename output folder out -> build.

The derived relations can be produced using the following operations over relations~\cite{alglave2016syntax}: % 3.1.6 Relations between events

\begin{itemize}
	\item \textit{the union} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,|\,r2}
	\item \textit{the intersection} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,\&\,r2}
	\item \textit{the difference} of two relations \texttt{r1} and \texttt{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}
\footnote{The sequence of two relations \texttt{r1} and \texttt{r2} is defined as the set of pairs \texttt{(x,y)} such that there exists an intervening \texttt{z}, such that $\mathtt{(x,z)} \in \mathtt{r1}$ and $\mathtt{(z,y)} \in \mathtt{r2}$}
	of two relations \texttt{r1} and \texttt{r2} is \texttt{r1;r2}
\end{itemize}

%The \textit{from-read order} \fr $\subset~\mathbb{R}~\times~\mathbb{W}$ that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).

%TODO: perephrase last sentence! alglave thesis, p. 36

@@

The semantics of a concurrent program is represented by the set of allowed executions.
A \textit{candidate execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location~\cite{alglave2014herding}.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph.

%An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}. 
%An execution to be checked on validity is called the \textit{candidate execution}.
An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$~\cite{alglave2010shared}.
As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program as a whole.
Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus-test Example~\ref{simple_wmm_x86} (the pictures are generated by the \texttt{herd7} tool, version 7.47). Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it. Since each thread performs single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location). Thus, there are only four possible executions defined by the choice of \rf-relation. The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models x86-TSO, Power, ARM, and others. However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over \fr$\cup$\po.
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

\section{The CAT language}

One common way to define a weak memory model in a systematic way in to formalise it in \textit{the CAT language} proposed in~\cite{alglave2016syntax}. The CAT language allows to axiomatically define new relations, new fences and restrictions over relations.

>>>>

%The \textit{from-read order} \fr $\subset~\mathbb{R}~\times~\mathbb{W}$ that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

@@

%The semantics of a concurrent program is represented as the set of allowed executions.
An \textit{execution} (trace, run) of a program is an ordered set of events defined by \po- and \rf-relations and set of final writes to a given memory location that is valid under certain memory model~\cite{alglave2014herding}. A \textit{candidate execution} is an execution that is not constrained by the memory model (i.e., the execution for that the validity is not checked). As it was shown in~\cite{wickerson2017automatically}, it is enough for memory models to constrain the executions independently instead of constraining the program at a whole.
%The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph. 
%An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}.

%An execution is uniquely defined by the set $\mathbb{X}$ of events have been executed in each thread (the \textit{control-flow} of a program), and the relations $\mathtt{rf}$ and $\mathtt{co}$~\cite{alglave2010shared}.
The Figure~\ref{simple_wmm_x86_pic} illustrates four possible candidate executions for the litmus test Example~\ref{simple_wmm_x86} (the pictures are generated by the \texttt{herd7} tool, version 7.47). Since there are no conditional jumps, the \po-relation is defined and we do not need to guess it. Since each thread performs single write followed by a single read, the \co-relation is also defined (it relates the initial write event with the write event to the same location). Thus, there are only four possible executions defined by the choice of \rf-relation. The candidate executions pictured in Figures~\ref{simple_wmm_x86_pic:sub1}--\ref{simple_wmm_x86_pic:sub3} are consistent both under strong memory model SC and under relaxed memory models x86-TSO, Power, ARM, and others. However, the execution shown in Figure~\ref{simple_wmm_x86_pic:sub3} is still consistent under relaxed-memory architectures, but it becomes inconsistent under SC architecture as it forbids cycles over \fr$\cup$\po.
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

\section{The \cat language}

Weak memory models can be described in the \cat language~\cite{alglave2016syntax}. It is the domain specific language for describing consistency properties of concurrent programs. The \cat language combines the expressive power of a functional language (it is inspired by OCaml and adopts its types, first-class functions, pattern matching and other features) with types, expressions and assertions specific for operating with relations and sets of executions.
% succinctly by constraining an abstraction of parallel program executions into a candidate execution and possibly extending this candidate execution with additional constraints on the execution environment.

%represents the functional language extended by the theory of  allows to axiomatically define derived relations, architecture-specific fences, and constraints over relations.

The definition of new relations  derived relations can be produced using the following operations over relations~\cite{alglave2016syntax}: % 3.1.6 Relations between events

\begin{itemize}
	\item \textit{the union} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,|\,r2}
	\item \textit{the intersection} of two relations \texttt{r1} and \texttt{r2} is \texttt{r1\,\&\,r2}
	\item \textit{the difference} of two relations \texttt{r1} and \texttt{r2} is $\mathtt{r1\,\backslash\,r2}$
	\item \textit{the sequence}\footnote{The sequence of two relations \texttt{r1} and \texttt{r2} is defined as the set of pairs \texttt{(x,y)} such that there exists an intervening \texttt{z}, such that $\mathtt{(x,z)} \in \mathtt{r1}$ and $\mathtt{(z,y)} \in \mathtt{r2}$}
	of two relations \texttt{r1} and \texttt{r2} is \texttt{r1;r2}
\end{itemize}
